Incremental Forward Reasoning for White-Box~Proof~Search

### 2025-02-18 paper.tex
partial eval changes

\renewcommand\UrlFont{\color{blue}\rmfamily}
\urlstyle{rm}
\def\orcid#1{{\href{http://orcid.org/#1}{\protect\raisebox{-1.25pt}{\protect\includegraphics{orcid.pdf}}}}}
\SetKw{KwIn}{in}
\SetKw{KwLet}{let}
\SetKw{KwYield}{yield}
\SetKwInOut{Input}{Input}
\SetKwInOut{Output}{Output}
\SetKwProg{Procedure}{Procedure}{}{}

@@

For each benchmark, we set up certain forward rules and a goal with certain hypotheses.
We then run the \texttt{saturate} tactic, which uses either naive or incremental forward reasoning to exhaustively apply the forward rules.
Performing only forward reasoning is the best-case scenario for the incremental algorithm since previous results can be fully reused.
When forward reasoning is combined with other rules, we would expect smaller gains, depending on how much reuse the non-forward rules allow.

Benchmark code and data are available in our supplement (TODO ref).
The results shown below were obtain on a Macbook Pro with an Apple~M2 Pro processor and \SI{32}{GB} of RAM, averaging over \textcolor{red}{N} (TODO) runs per benchmark.

\paragraph{Transitivity.}
Our first benchmark is a scaled-up variant of the transitivity example from Sec.~\ref{sec:incremental}.
Given some relation $≺$ on natural numbers, we register the lemma $\All{x~y}{x ≺ y → y ≺ z → x ≺ z}$ as the sole forward rule and set up a goal with hypotheses $h₁ : a ≺ a + 1, \dots, hₙ : a + n - 1 ≺ a + n$ (TODO what is $a$?).
Saturating this goal adds $n(n-1)/2$ hypotheses to the context.

Instead of the natural number type of Lean's standard library, which has special support in the compiler and kernel, we use a custom type of Peano naturals.
Unifying two concrete Peano natural numbers $n$ and $m$ involves comparing $\min(n, m)$ constructors.
We can therefore use moderately large numbers to simulate the bigger unification problems that Aesop encounters, for example, in Mathlib's category theory development.

\begin{figure}
  \centering
  \begin{subfigure}{.5\textwidth}
    \centering
    \begin{tikzpicture}[scale=0.75]
      \begin{axis}[
        xlabel={Number of hypotheses},
        ylabel={Time in ms},
        xmin=1, xmax=16,
        ymin=2^3, ymax=2^13,
        xtick={1,2,4,8,16},
        ytick={2^3,2^5,2^7,2^9,2^11,2^13},
        xmode=log,
        ymode=log,
        log basis x=2,
        log basis y=2,
        legend pos=north west,
        ymajorgrids=true,
        grid style=dashed,
        legend image post style={mark=},
        ]
        \addplot[
          color=orange,
          mark=square,
          ]
          coordinates {
            (1, 7.663472) (2, 9.354750) (4, 15.768194) (8, 104.921791) (16, 1882.683111)
          };
        \addlegendentry{naive}
        \addplot[
          color=blue,
          mark=square,
          ]
          coordinates {
            (1, 10.308541) (2, 10.704638) (4, 16.274305) (8, 38.800833) (16, 179.697681)
          };
        \addlegendentry{incremental}
        \addplot[
          color=orange,
          mark=o,
          ]
          coordinates {
            (1, 10.591138) (2, 23.881333) (4, 142.013389) (8, 1676.148375) (16, 24327.359417)
          };
        \addplot[
          color=blue,
          mark=o,
          ]
          coordinates {
            (1, 13.974694) (2, 26.216833) (4, 68.953444) (8, 245.601028) (16, 1028.670680)
          };
      \end{axis}
    \end{tikzpicture}
    \caption{Transitivity benchmark}%
    \label{fig:trans}
  \end{subfigure}%
  \begin{subfigure}{.5\textwidth}
    \centering
    \begin{tikzpicture}[scale=0.75]
      \begin{axis}[
        xlabel={Number of Rules},
        %ylabel={Time in ms},
        xmin=1, xmax=32,
        ymin=2^3, ymax=1700,
        ymode=log,xmode=log,
        log basis x=2,
        log basis y=2,
        legend pos=north west,
        ymajorgrids=true,
        grid style=dashed,
        legend image post style={mark=},
        ]

        \addplot[
          color=orange,
          mark=square,
          ]
          coordinates {
            (1, 10.399958) (2, 16.426541) (4, 43.461403) (8, 205.595541) (16, 1238.090889) (32, 8927.857819)
          };
        \addlegendentry{naive}
        \addplot[
          color=blue,
          mark=square,
          ]
          coordinates {
            (1, 11.231875) (2, 12.879500) (4, 16.851930) (8, 24.666485) (16, 42.054055) (32, 76.369778)
          };
        \addlegendentry{incremental}
        \addplot[
          color=orange,
          mark=o,
          ]
          coordinates {
            (1, 19.405222) (2, 33.487472) (4, 80.107597) (8, 284.954791) (16, 1536.089624) (32, 10721.729861)
          };
        \addplot[
          color=blue,
          mark=o,
          ]
          coordinates {
            (1, 25.892069) (2, 40.031930) (4, 69.446902) (8, 130.640986) (16, 260.653917) (32, 481.827902)
          };
      \end{axis}
    \end{tikzpicture}
    % graph \includegraphics[width=.4\linewidth]{image1}
    \caption{Independence benchmark}%
    \label{fig:indep}
  \end{subfigure}
  \caption{The above results depict the behavior of both algorithms in a context where there is a lot of work that can be reused (Transitivity) and no work to be reused (Independence). The lines with square entry points represent problems where unification is easy. The lines with circle entry points are the same tests with hard unification problems.}%
  \label{fig:benchmark}
  \end{figure}

The results in Fig.~\ref{fig:trans} are unsurprising.
The naive algorithm goes over the context, containing initially $n$ hypotheses, and compares all hypotheses pairwise.
This is repeated approximately for each of the $n(n-1)/2$ added hypotheses, so the overall complexity is $O(n^4)$.
The incremental algorithm considers each hypothesis only once.
As the hypotheses get matched to both slots, it then compares each hypothesis with at most $n$ other hypotheses.
This yields a complexity of $O(n^2)$.

@@

Given a set of propositions $\mathcal{P} := \{P_{i,j} | 1 \leq i \leq n \text{ and } 1 \leq j \leq 6\}$ and $\mathcal{Q} :=\{Q_{i} | 1 \leq i \leq n \}$, in this test, we consider the $n$ following rules $r_i : P_{i,1}\to \dots P_{i,6} \to Q_i$.
The procedures are then run on a context containing exactly $\mathcal{P}$.
In this case, we sketch the behaviour of the stateless algorithm:
It pick a first hypothesis, $h_{i,j} : P_{i,j}$.
If $j < 6$, the index returns nothing and it selects another hypothesis.
For $h_{i,6} : P_{i,6}$, the rule $r_i$ is selected, and it then needs to try and fill its remaining five slot with the context's $6n$ hypotheses which it tries all.
This gives a complexity of about $O(n^2)$.

>>>>

\renewcommand\UrlFont{\color{blue}\rmfamily}
\urlstyle{rm}
\def\orcid#1{{\href{http://orcid.org/#1}{\protect\raisebox{-1.25pt}{\protect\includegraphics{orcid.pdf}}}}}
\def\showpgfcircle{\tikz[baseline=-0.7ex]\node[mark size=0.7ex]
{\pgfuseplotmark{o}};}
\def\showpgfsquare{\tikz[baseline=-0.7ex]\node[mark size=0.7ex]
{\pgfuseplotmark{square}};}
\SetKw{KwIn}{in}
\SetKw{KwLet}{let}
\SetKw{KwYield}{yield}
\SetKwInOut{Input}{Input}
\SetKwInOut{Output}{Output}
\SetKwProg{Procedure}{Procedure}{}{}

@@

For each benchmark, we set up certain forward rules and a goal with handpicked hypotheses.
We then run the \texttt{saturate} tactic, which uses either naive or incremental forward reasoning to exhaustively apply the forward rules.
Performing only forward reasoning is the best-case scenario for the incremental algorithm since previous results can be fully reused.
When forward reasoning is combined with other rules, we would expect smaller gains, depending on how much reuse the non-forward rules allow.

In what follows, instead of the natural number type of Lean's standard library, which has special support in the compiler and kernel, we use a custom type of Peano naturals.
Unifying two concrete Peano natural numbers $n$ and $m$ involves comparing $\min(n, m)$ constructors.
We can therefore use moderately large numbers to simulate the bigger unification problems that Aesop encounters, for example, in Mathlib's category theory development.

Benchmark code and data are available in our supplement (TODO ref).
The results shown below were obtained on a MacBook Pro with an Apple~M2 Pro processor and \SI{32}{GB} of RAM.
The results are an average of \textcolor{red}{N} (TODO) runs per benchmark.

\paragraph{Transitivity.}
Our first benchmark is a scaled-up variant of the transitivity example from Sec.~\ref{sec:incremental}.
Given some relation $≺$ on natural numbers, we register the lemma $\All{x~y}{x ≺ y → y ≺ z → x ≺ z}$ as the sole forward rule and set up a goal with hypotheses $h₁ : a ≺ a + 1, \dots, hₙ : a + n - 1 ≺ a + n$. The test is run for $a = 0$, represented with a square \protect\showpgfsquare\, and $a = 100$, represented with a circle \protect\showpgfcircle\.
Saturating this goal adds $n(n-1)/2$ hypotheses to the context.

\begin{figure}
  \centering
  \begin{subfigure}{.5\textwidth}
    \centering
    \begin{tikzpicture}[scale=0.75]
      \begin{axis}[
        xlabel={Number of hypotheses},
        ylabel={Time in ms},
        xmin=1, xmax=16,
        ymin=4, ymax=2^15,
        %xtick={1,2,4,8,16},
        %ytick={2^3,2^5,2^7,2^9,2^11,2^13},
        xmode=log,
        ymode=log,
        log basis x=2,
        log basis y=2,
        legend pos=north west,
        ymajorgrids=true,
        grid style=dashed,
        legend image post style={mark=},
        ]
        \addplot[
          color=orange,
          mark=square,
          ]
          coordinates {
            (1, 7.663472) (2, 9.354750) (4, 15.768194) (8, 104.921791) (16, 1882.683111) % (32, 1175.888750) 
          };
        \addlegendentry{naive}
        \addplot[
          color=blue,
          mark=square,
          ]
          coordinates {
            (1, 10.308541) (2, 10.704638) (4, 16.274305) (8, 38.800833) (16, 179.697681) % (32, 45220.919667) 
          };
        \addlegendentry{incremental}
        \addplot[
          color=orange,
          mark=o,
          ]
          coordinates {
            (1, 10.591138) (2, 23.881333) (4, 142.013389) (8, 1676.148375) (16, 24327.359417) %(32, 392049.849708) 
          };
        \addplot[
          color=blue,
          mark=o,
          ]
          coordinates {
            (1, 13.974694) (2, 26.216833) (4, 68.953444) (8, 245.601028) (16, 1028.670680) %(32, 4388.917500) 
          };
      \end{axis}
    \end{tikzpicture}
    \caption{Transitivity benchmark}%
    \label{fig:trans}
  \end{subfigure}%
  \begin{subfigure}{.5\textwidth}
    \centering
    \begin{tikzpicture}[scale=0.75]
      \begin{axis}[
        xlabel={Number of Rules},
        %ylabel={Time in ms},
        xmin=1, xmax=32,
        ymin=4, ymax=2^15,
        ymode=log,xmode=log,
        log basis x=2,
        log basis y=2,
        legend pos=north west,
        ymajorgrids=true,
        grid style=dashed,
        legend image post style={mark=},
        ]

        \addplot[
          color=orange,
          mark=square,
          ]
          coordinates {
            (1, 10.399958) (2, 16.426541) (4, 43.461403) (8, 205.595541) (16, 1238.090889) (32, 8927.857819)
          };
        \addlegendentry{naive}
        \addplot[
          color=blue,
          mark=square,
          ]
          coordinates {
            (1, 11.231875) (2, 12.879500) (4, 16.851930) (8, 24.666485) (16, 42.054055) (32, 76.369778)
          };
        \addlegendentry{incremental}
        \addplot[
          color=orange,
          mark=o,
          ]
          coordinates {
            (1, 19.405222) (2, 33.487472) (4, 80.107597) (8, 284.954791) (16, 1536.089624) (32, 10721.729861)
          };
        \addplot[
          color=blue,
          mark=o,
          ]
          coordinates {
            (1, 25.892069) (2, 40.031930) (4, 69.446902) (8, 130.640986) (16, 260.653917) (32, 481.827902)
          };
      \end{axis}
    \end{tikzpicture}
    % graph \includegraphics[width=.4\linewidth]{image1}
    \caption{Independence benchmark}%
    \label{fig:indep}
  \end{subfigure}
  \caption{The above results depict the behavior of both algorithms in a context where a lot of work can be reused (Transitivity) and in another where no work can be reused (Independence). 
  In both benchmarks, results denoted with a square \protect\showpgfsquare\ come from a context where the argument of the predicates is small ($a = 0$), leading to easy unification problems; results denoted with a circle \protect\showpgfcircle\ come from a context where the unification problem is hard ($a = 100$).}%
  \label{fig:benchmark}
  \end{figure}

We see in Fig.~\ref{fig:trans} that the incremental algorithm has a sizable advantage over the naive algorithm.
An important detail is that the scales are logarithmic
These results are unsurprising:
The naive algorithm goes over the context, containing initially $n$ hypotheses, and compares all hypotheses pairwise.
This is repeated approximately for each of the $n(n-1)/2$ added hypotheses, so the overall complexity is $O(n^4)$.
The incremental algorithm considers each hypothesis only once.
As the hypotheses get matched to both slots, it then compares each hypothesis with at most $n$ other hypotheses.
This yields a complexity of $O(n^2)$.

@@

Given a set of predicates $\mathcal{P} := \{P_{i,j} (x) | 1 \leq i \leq n \text{ and } 1 \leq j \leq 6\}$ and $\mathcal{Q} :=\{Q_{i}(x) | 1 \leq i \leq n \}$, in this test, we consider the $n$ following rules $r_i : \forall x, P_{i,1}(x)\to \dots P_{i,6}(x) \to Q_i(x)$.
The procedures are then run on a context containing exactly $\mathcal{P}$.
In this case, we sketch the behaviour of the stateless algorithm:
It pick a first hypothesis, $h_{i,j} : P_{i,j}(a)$.
If $j < 6$, the index returns nothing and it selects another hypothesis.
For $h_{i,6} : P_{i,6}(a)$, the rule $r_i$ is selected, and it then needs to try and fill its remaining five slot with the context's $6n$ hypotheses which it tries all.
This gives a complexity of about $O(n^2)$.
